---
title: "Interactive analysis"
subtitle: Statistical Programming using `R`
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      results = 'markdown')
library(markdown)
inclure <- function(fichier) {
    name = tail(strsplit(fichier, "/")[[1]], 1)
    tags$div(
        tags$ul(tags$li(tags$a(tags$code(name), href = fichier))),
        tags$pre(tags$code(includeText(fichier)),
             class = "r")
    )
}
```

We present in this section the `shiny` package, which allows us to create web application (local to a computer, local into an inside network, or online).

```{r, echo=TRUE}
library(shiny)
```

We see in the previous sessions that it is a diffuclt task to analysis deeply some datasets, when we produce a lot of tables and graphics. A good way to tackle this problem is to create a (simple) application, which allows us to interact with data and methods. 

## Web application

To create a web application with `shiny`, you have to create a specific folder with specifics files:

- *one file app*: file named `app.R`
- *two files app*: files names `server.R` and `ui.R` 

I prefer the second choice to separate the *user interface* side (in `ui.R`) and the *server* side (in `server.R`). 

Since we want an interactive application, we have to manage both sides of it:

- **UI**: we explain here what is the format of our application (title, panels, ...), and the possible interactions
    - **inputs**: parameters the user can define
    - **outputs**: tables and graphics produce by the server
- **Server**: we detail here how `R` produces the **outputs**, from data and **inputs** 

In the following subsection are presented the contents of the two files. You can copy this to test inside **RStudio**.

### First application

We create a folder, names for instance `web-application`, with two files:

```{r}
inclure("stat-prog-R/web-application-1/ui.R")
inclure("stat-prog-R/web-application-1/server.R")
```

In **RStudio**, when a file named `ui.R` or `server.R` or `app.R` is open, you get a button names *Run App*. If you click on them, you must have a new window with only the main title.

The funtion `shinyUI()` define the **UI** section of an application. In this, we create a `fluidPage()`, which can contain many *panel*, as we see later.

The function `shinyServer()` permits us to define a function taking `input` and `output` parameters defined in `ui.R`, to explain how **outputs** are created from data and **inputs**.

If you want to run directly the application, write the following code into `R` console.

```{r, eval=FALSE, echo=TRUE}
runUrl("http://fxjollois.github.io/stat-prog-R/web-application-1.zip")
```


### Adding plot and table

We add in the preceding application a plot and a table, without any interaction at this moment. So, we have to define the outputs in `ui.R`, with functions names `plotOutput()` for any plot, and `tableOutput()` for tables. These functions must have a parameter, which is the name of the output. 

Now, in `server.R`, we define the plot and the table to produce. We use `iris` dataset, with functions `renderPlot()` for any plot, and `renderTable()` for tables.

To enhance the presentation of the application, we use a `sidebarLayout` which divides the window into two sections:

- `sidebarPanel()`: usually all the controls (interactions)
- `mainPanel()`: the outputs generated from choice made by users

And in the `server.R`, we compute *PCA* on the data before the definition of the server operation.

```{r}
inclure("stat-prog-R/web-application-2/ui.R")
inclure("stat-prog-R/web-application-2/server.R")
```

If you want to run directly the application, write the following code into `R` console.

```{r, eval=FALSE, echo=TRUE}
runUrl("http://fxjollois.github.io/stat-prog-R/web-application-2.zip")
```

### Adding some interaction

Now, we add interactions in the application, then the user (you or someone else if you publish your app) can interact with data. First, we add two inputs into `ui.R`:

- One drop down list to chose the statistic computing in the table, with `selectInput()` function ;
- Two groups of radio buttons to chose the dimension to plot, with `radioButtons()`.

These two functions take at least three parameters:

- a name, to be use in `server.R`
- a title, placed above the drop down menu
- a list of possible choices

For `radioButtons()`, we add the parameter `inline=TRUE` to have all the button on one line.

Second, we use theses inputs as object of `input` parameter in the server function, to use the user choice in the produced outputs. The value inside these inputs are *characters*.

```{r}
inclure("stat-prog-R/web-application-3/ui.R")
inclure("stat-prog-R/web-application-3/server.R")
```

If you want to run directly the application, write the following code into `R` console.

```{r, eval=FALSE, echo=TRUE}
runUrl("http://fxjollois.github.io/stat-prog-R/web-application-3.zip")
```

### Add updates and dependencies

At the end, we add in our application three upgrades:

1. we update possible choices in the radio buttons according to the number of dimensions of our PCA result
    - `ui.R`: the only choice is `0` at the beginning of the application
    - `server.R`: we indicate the new choices (between 1 and the number of dimension in PCA) in a `observe()` function, with `updateRadioButtons()`
2. we add dependency between `dimY` and `dimX` inputs, to change `dimY` value according to a new value of `dimX`
    - `server.R`: we modify the selected value in `dimY` (with `updateRadioButtons()`) to the first value different from `dimX`
3. we create a *reactive* object to get the subset of PCA dimensions
    - `server.R`: 
        - create a function with `reactive()` returning the subset with the two choosen dimensions
        - get the data by using the created function in the `renderPlot()`

We add some controls in the outputs to avoid error during the execution of the application.

```{r}
inclure("stat-prog-R/web-application-4/ui.R")
inclure("stat-prog-R/web-application-4/server.R")
```

If you want to run directly the application, write the following code into `R` console.

```{r, eval=FALSE, echo=TRUE}
runUrl("http://fxjollois.github.io/stat-prog-R/web-application-4.zip")
```

### User interface

We can modify the user interface in many different ways with `shiny` package. It exists also some other packages allowing us to personnalize applications (`shinydashboard`, `shinythemes`, ...)

## Some work

Use the `shiny` package to create an application able to help us in the clustering process of our data.
