---
title: "Introduction to Data Analysis & Data Visualization"
subtitle: Statistical Programming using `R`
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = TRUE)
```

All the following code can be executed into `R` console, without any additional *packages* installation or *data* laoding, except when indicated.

## Introduction to `R` language

### Some useful things

#### Basic commands

```{r}
1 + 2
a = 1
a + 2
c(1, 2)
c(1, 2) + 3
1:5
seq(1, 5)
seq(1, 5, by = .5)
seq(1, 5, length = 9)
rep(1, 5)
rep(c(1, 3), times = 5)
rep(c(1, 3), each = 5)
rep(c(1, 3), times = 2, each = 5)
rep(c(1, 3), times = 2, each = 5, length = 11)
ifelse(1 > 2, "sup", "inf")
for(v in 1:5) print(v**2)
sample(1:10)
paste("good", "morning")
paste("good", "morning", sep = "-")
paste0("good", "morning")
paste("good", c("morning", "afternoon"))
```


#### Working directory

```{r, eval = FALSE}
# this code is not executed
getwd()
setwd("path/to/your/working/directory")
getwd() # if no change, problem with previous line
```

#### Packages

Packages directory

```{r, eval = FALSE}
# this code is not executed
.libPaths()
.libPaths("path/to/your/packages/directory")
.libPaths() # if no change, problem with previous line
```

Package installation

```{r, eval = FALSE}
# this code is not executed
install.packages("packagename") # one or more packages name
library(package) # without any quote
```

#### Functions

```{r}
f <- function(a, b) {
    return(a ** b)
}
f(2, 5)
f(a = 2, b = 5)
f(a = 2, 5)
f(b = 2, a = 5)
f(b = 2, 5)
```

```{r}
f <- function(a, b = 1) {
    return(a ** b)
}
f(2, 5)
f(2)
```

```{r}
f <- function(a, b = 1, ...) {
    s = sum(a, ...)
    return(s ** b)
}
f(2, 5)
f(c(1, 2, 3), 5)
f(c(1, NA, 3), 5)
sum(c(1, NA, 3))
sum(c(1, NA, 3), na.rm = TRUE)
f(c(1, NA, 3), 5, na.rm = TRUE)
```

#### Import data from text files

We present here different files, with specific needs to correctly import data into `R`. We use the `read.table()` command, with options.

##### [heart.txt](donnees/heart.txt)

- Tabulated separated variables (default value)
- attriutes names in the first line (`header = T`)

```{r}
heart = read.table("donnees/heart.txt", header = T)
head(heart)
```

##### [hepatitis.TXT](donnees/hepatitis.TXT)

- Tabulated separated variables
- attributes names in the first line (`header = T`)
- missing value indicated with `"?"` (`na.strings = "?"`)

```{r}
hep = read.table("donnees/hepatitis.TXT", header = T, na.strings = "?")
head(hep)
```

##### [Iris_bis.txx](donnees/Iris_bis.txt)

- delimited variables with `"%"` (`sep = "%"`)
- attributes names in the first line (`header = T`)
- specific decimal separator (`dec = "."`)

```{r}
ir = read.table("donnees/Iris_bis.txt", header = T, sep = "%", dec = ",")
```

##### [Detroit_homicide.txt](donnees/Detroit_homicide.txt)

- attributes names in the first line (`header = T`)
- 35 first rows useless (`skip = 35`)

```{r}
dh = read.table("donnees/Detroit_homicide.txt", skip = 35, header = T)
head(dh)
```


### Working with `data.frame`

About `data.frame` (basic manipulation)

```{r}
head(iris)
class(iris)
names(iris)
dim(iris)
nrow(iris)
ncol(iris)
str(iris)
```

For illustration, we extract 20 random rows from `iris`.

```{r}
iris = head(iris[sample(seq(nrow(iris))),], 20)
```

Get one column from a `data.frame`

```{r}
iris$Sepal.Length
iris[,1]
iris[,"Sepal.Length"]
iris[1]
iris["Sepal.Length"]
```

Get more than one column 

```{r}
iris[,c(1,3)]
iris[,c("Sepal.Length", "Petal.Length")]
iris[c(1,3)]
iris[c("Sepal.Length", "Petal.Length")]
iris[-c(2, 4, 5)]
```

Get one row from a `data.frame`

```{r}
iris[1,]
iris[c(1,3),]
iris[iris$Sepal.Length > 6,]
```

Data types

```{r}
class(names(iris))
class(iris$Sepal.Length)
class(iris$Species)
head(iris$Species)
```

Group two `data.frame` together

```{r}
cbind(iris[,1], iris[,4:5])
rbind(iris[1:3,], iris[8:10,])
```


```{r, include = FALSE}
rm(iris)
```

### Univariate statistics

We now work on the complete `iris` dataset.

#### Quantitative 

##### Numerical statistics

```{r}
mean(iris$Sepal.Length)
sd(iris$Sepal.Length)
var(iris$Sepal.Length)
median(iris$Sepal.Length)
min(iris$Sepal.Length)
max(iris$Sepal.Length)
range(iris$Sepal.Length)
quantile(iris$Sepal.Length)
quantile(iris$Sepal.Length, .1)
quantile(iris$Sepal.Length, c(.01, .1, .9, .99))
summary(iris$Sepal.Length)
shapiro.test(iris$Sepal.Length)
```

##### Basic graphics

```{r}
hist(iris$Sepal.Length)
boxplot(iris$Sepal.Length)
qqnorm(iris$Sepal.Length)
qqline(iris$Sepal.Length)
plot(ecdf(iris$Sepal.Length))
```


#### Qualitative

`factor` 

##### Numerical statistics

```{r}
t = table(iris$Species)
prop.table(t)
chisq.test(t)
chisq.test(t, p = c(.2, .3, .5))
```

##### Basic graphics

```{r}
barplot(table(iris$Species))
plot(iris$Species) # work only because Species is declared as factor
pie(table(iris$Species))
```


### Bivariate statistics

#### Two quantitatives

##### Numerical

```{r}
cov(iris$Sepal.Length, iris$Sepal.Width)
cor(iris$Sepal.Length, iris$Sepal.Width)
cor.test(iris$Sepal.Length, iris$Sepal.Width)
m = lm(Sepal.Length ~ Sepal.Width, data = iris)
m
summary(m)
```

##### Graphics

```{r}
plot(iris$Sepal.Length, iris$Sepal.Width)
plot(iris$Sepal.Length ~ iris$Sepal.Width)
plot(Sepal.Length ~ Sepal.Width, data = iris)
abline(m, col = "red")
```


#### Two qualitatives

Create a new ordinal attribute from `Sepal.Length`, with 5 values for instance.

```{r}
iris$Sepal.Length.ord = cut(iris$Sepal.Length, 5)
table(iris$Sepal.Length.ord)
diff(range(iris$Sepal.Length)) / 5
```


##### Numerical statistics

```{r}
t = table(iris$Species, iris$Sepal.Length.ord)
t
prop.table(t)
prop.table(t, margin = 1)
prop.table(t, margin = 2)
chisq.test(t)
```

##### Graphics

```{r}
barplot(t)
barplot(t, beside = TRUE)
barplot(prop.table(t, margin = 2))
assocplot(t)
mosaicplot(t)
```

#### One quantitative vs. one qualitative

##### Numerical statistics

```{r}
tapply(iris$Sepal.Length, iris$Species, mean)
tapply(iris$Sepal.Length, iris$Species, summary)
anova(lm(Sepal.Length ~ Species, data = iris))
m = aov(Sepal.Length ~ Species, data = iris)
m
summary(m)
```

##### Graphics

```{r}
boxplot(Sepal.Length ~ Species, data = iris)
```

```{r}
par(mfrow = c(3, 1), mar = c(2, 2, 0, 0) + .1)
x = iris$Sepal.Length
for (v in levels(iris$Species)) {
    hist(x[iris$Species == v], 
         freq = FALSE,
         xlim = range(x),
         breaks = seq(min(x), max(x), length = 10),
         main = NULL)
}
```


### Multivariate statistics (basic)

on `data.frame`

```{r}
summary(iris)
apply(iris[1:4], 2, mean)
apply(iris[1:4], 2, summary)
apply(iris[1:4], 2, tapply, iris$Species, mean)
cor(iris[1:4])
```

```{r}
plot(Sepal.Length ~ Sepal.Width, data = iris, 
     col = rainbow(3)[iris$Species], pch = 19)
legend("topright", 
       pch = 19, col = rainbow(3), 
       legend = levels(iris$Species))
```

```{r}
pairs(iris[-5])
pairs(iris[-5], col = rainbow(3)[iris$Species], pch = 19, upper.panel = NULL)
```


### Advanced manipulation, using basic `R` code

#### `subset()`

```{r}
subset(iris, subset = Species == "setosa")
subset(iris, subset = Species == "setosa" & Sepal.Length > 5.5)
subset(iris, select = c(Sepal.Length, Species))
subset(iris, 
       subset = Species == "setosa" & Sepal.Length > 5.5, 
       select = c(Sepal.Length, Species))
```

#### `transform()`

```{r}
transform(iris, Sepal.Length.ord = cut(Sepal.Length, 5))
```


#### `aggregate()`

```{r}
aggregate(Sepal.Length ~ 0, iris, mean)
aggregate(Sepal.Length ~ 0, iris, length)
aggregate(Sepal.Length ~ Species, iris, mean)
aggregate(Sepal.Length ~ Species, iris, length)
aggregate(cbind(Sepal.Length, Sepal.Width) ~ Species, iris, mean)
aggregate(. ~ Species, iris, mean)
aggregate(Sepal.Length ~ Species, iris, function(v) {
    return(c(mean = mean(v), sd = sd(v)))
    })
```


#### `setNames()`

```{r}
setNames(aggregate(Sepal.Length ~ Species, iris, mean),
         c("Species", "Sepal Length mean"))
```

#### `merge()`

```{r}
a = aggregate(Sepal.Length ~ Species, iris, mean)
a = setNames(a, c("Species", "Sepal Length mean"))
b = aggregate(Sepal.Width ~ Species, iris, min)
b = setNames(b, c("Species", "Sepal Width min"))
merge(a, b)
```

```{r}
a = aggregate(Sepal.Length ~ Species, iris, mean)
b = aggregate(Sepal.Width ~ Species, iris, min)
merge(a, b, by = "Species")
```

## Data visualization

In the following code, we need two packages:

- [`ggplot2`]()
- [`FactoMineR`]()

### library `ggplot2`

```{r}
library(ggplot2)
```


Two functions: `qplot()` and `ggplot()`. The first one is a simplier version of the second one. We focus here on the `ggplot()`.

#### Univariate 

##### Quantitative

```{r}
ggplot(iris, aes(Sepal.Length)) + geom_histogram()
ggplot(iris, aes(Sepal.Length)) + geom_density()
ggplot(iris, aes(Sepal.Length)) + 
    geom_histogram(aes(y = ..density..)) + 
    geom_density()
ggplot(iris, aes(sample = Sepal.Length)) + geom_qq()
ggplot(iris, aes(0, Sepal.Length)) + geom_boxplot()
ggplot(iris, aes(Sepal.Length)) + stat_ecdf()
```

##### Qualitative

```{r}
ggplot(iris, aes(Sepal.Length.ord)) + geom_bar()
ggplot(iris, aes(Sepal.Length.ord)) + 
    geom_bar(aes(y = (..count../(sum(..count..))))) 
library(scales)
ggplot(iris, aes(Sepal.Length.ord)) + 
    geom_bar(aes(y = (..count../(sum(..count..))))) +
    scale_y_continuous(labels = percent) 
ggplot(iris, aes("", fill = Sepal.Length.ord)) + 
    geom_bar(aes(y = (..count..)/sum(..count..))) +
    scale_y_continuous(labels = percent)
ggplot(iris, aes("", fill = Sepal.Length.ord)) + 
    geom_bar(aes(y = (..count..)/sum(..count..)), width = 1) +
    scale_y_continuous(labels = percent) +
    coord_polar(theta = "y")
```

#### Bivariate

##### Two Quantitatives

```{r}
ggplot(iris, aes(Sepal.Length, Sepal.Width)) + geom_point()
ggplot(iris, aes(Sepal.Length, Sepal.Width)) + geom_rug()
ggplot(iris, aes(Sepal.Length, Sepal.Width)) + geom_smooth()
ggplot(iris, aes(Sepal.Length, Sepal.Width)) + geom_smooth(method = "lm")
ggplot(iris, aes(Sepal.Length, Sepal.Width)) + 
    geom_point() + 
    geom_rug() + 
    geom_smooth(method = "lm")
ggplot(iris, aes(Sepal.Length, Sepal.Width)) + geom_bin2d(bins = 10)
```

##### Two qualitatives

```{r}
ggplot(iris, aes(Sepal.Length.ord)) + geom_bar() + facet_wrap(~ Species)
ggplot(iris, aes(Sepal.Length.ord, fill = Species)) + geom_bar()
ggplot(iris, aes(Sepal.Length.ord, fill = Species)) + 
    geom_bar(position = "fill") +
    scale_y_continuous(labels = percent)
ggplot(iris, aes("", fill = Sepal.Length.ord)) + 
    geom_bar(aes(y = (..count..)/sum(..count..)), width = 1) +
    scale_y_continuous(labels = percent) +
    coord_polar(theta = "y") + 
    facet_wrap(~ Species)
ggplot(iris, aes(Species, Sepal.Length.ord)) + geom_bin2d()
```

##### Qualitative vs. Quantitative

```{r}
ggplot(iris, aes(Sepal.Length)) + geom_histogram() + facet_grid(Species ~.)
ggplot(iris, aes(Sepal.Length, fill = Species)) + 
    geom_histogram() + facet_grid(Species ~.)
ggplot(iris, aes(Sepal.Length, col = Species)) + geom_density()
ggplot(iris, aes(Species, Sepal.Length)) + geom_boxplot()
ggplot(iris, aes(Species, Sepal.Length, fill = Species)) + geom_boxplot()
ggplot(iris, aes(Sepal.Length, col = Species)) + stat_ecdf()
```

#### Multivariate

Some graphics with more than 2 variables

```{r}
ggplot(iris, aes(Petal.Length, Petal.Width, col = Species)) + 
    geom_point()
ggplot(iris, aes(Petal.Length, Petal.Width, col = Species)) + 
    geom_point() + facet_wrap(~ Sepal.Length.ord)
ggplot(iris, aes(Petal.Length, Petal.Width, col = Species)) + 
    geom_point() + facet_grid(Species ~ Sepal.Length.ord)
```

### library `FactoMineR`

For data analysis, such as *Principle Components Analysis* (**PCA**)

```{r}
library(FactoMineR)
```

```{r}
pca = PCA(iris, quali.sup = c(5, 6)) # since we add Sepal.Length.ord to iris data
summary(pca)
dimdesc(pca)
pca$eig
plot(pca$eig[,3], type = "b")
eig = data.frame(comp = rownames(pca$eig),
                 setNames(pca$eig, c("eigenvalue", "percent", "cum.percent")))
ggplot(eig) +
    geom_bar(aes(comp, percent), stat = "identity") +
    stat_summary(aes(comp, cum.percent, group = 1), 
                 fun.y = sum, geom = "line")
plotellipses(pca)
plot(pca, choix = "ind", habillage = 5, invisible = "quali", label = "none")
ggplot(cbind(iris, pca$ind$coord), aes(Dim.1, Dim.2, col = Species)) +
    geom_point() +
    stat_ellipse()
```



## Some work

Download the two following files :

- [pendigits.tra](donnees/pendigits.tra)
- [pendigits.tes](donnees/pendigits.tes)

Data description can be found [here](http://archive.ics.uci.edu/ml/datasets/Pen-Based+Recognition+of+Handwritten+Digits): each line represent a draw of a digit between 0 and 9, with 8 points ($(x,y)$ coordinates for each point) and the number drawn.

```{r, echo = FALSE}
pen.tra = read.table("donnees/pendigits.tra", sep = ",")
pen.tes = read.table("donnees/pendigits.tes", sep = ",")
pen = rbind(pen.tra, pen.tes)
names(pen) = c(paste0(c("X", "Y"), rep(1:8, each = 2)), "digit")
```

1. Import the two files, and group them into one `data.frame`
    - `data.frame` dimensions: `r nrow(pen)` rows and `r ncol(pen)` columns
    - column names: `r names(pen)`
1. Describe (quickly) all attributes
    - one by one first
    - coordinates with respect to `digit`
1. Represent the first `"0"` drawn as following
```{r, echo = FALSE}
first.0 = subset(pen, digit == 0)[1,]
x = unlist(first.0[seq(1,15, by = 2)])
y = unlist(first.0[seq(2,16, by = 2)])
par(mar = c(0, 0, 0, 0) + .1)
plot(x, y, type = "b", pch = " ", axes = FALSE, col = "gray70")
text(x, y, 1:8)
# d = data.frame(x = unlist(first.0[seq(1,15, by = 2)]), 
#                y = unlist(first.0[seq(2,16, by = 2)]),
#                n = 1:8)
# ggplot(d, aes(x, y)) +
#     geom_path(col = "gray70") +
#     geom_text(aes(label = n)) +
#     theme_void()
```
1. Use the preceding code to create a function, with
    - Parameters:
        - a vector with the 16 values of each point (if it is a `data.frame`, transform it as simple vector)
        - a numeric value (with `NULL` value by default), denoting the number to drawn (possibly not pass)
        - a boolean vector (with `FALSE` value by default), indicating if the point number has to be added to the graph
        - a boolean vector (with `FALSE` value by default), indicating if the plot is to added to the previous one
        - a specific color (with `"black"` value by default)
    - Output: 
        - line representing the drawn
        - number drawn as title (if known)
        - point number (if wanted)
```{r, echo = FALSE}
drawn <- function(v, n = NULL, point = FALSE, add = FALSE, color = "black") {
    if (is.data.frame(v))
        v = unlist(v)
    x = v[seq(1, 15, by = 2)]
    y = v[seq(2, 16, by = 2)]
    if (!add) {
        opar = par(mar = c(0, 0, 2, 0) + .1)
        plot(x, y, 
             xlim = c(-5, 105), ylim = c(-5, 105),
             axes = FALSE,
             type = ifelse(point, "b", "l"), 
             pch = " ", 
             col = color,
             main = n)
        if (point) text(x, y, 1:8)
        par(opar)
    } else {
        opar = par(mar = c(0, 0, 2, 0) + .1)
        lines(x, y, 
              xlim = c(-5, 105), ylim = c(-5, 105),
              type = ifelse(point, "b", "l"), 
              pch = " ", 
              col = color)
        par(opar)
    }
}
drawn(first.0)
drawn(first.0, n = 0)
drawn(first.0, point = TRUE)
drawn(first.0)
first.1 = subset(pen, digit == 1)[1,]
drawn(first.1, add = TRUE, col = "red")
```
1. Use the function to represent, in one graph, each first drawn for each number
```{r, echo = FALSE}
par(mfrow = c(2, 5))
for (i in 0:9) {
    s = subset(pen, digit == i)[1,]
    drawn(s[-16], s[17])
}
```
1. Compute the mean drawn for each number and represent it, with each drawn in (very) light gray
```{r, echo = FALSE}
par(mfrow = c(2, 5))
for (i in 0:9) {
    s = subset(pen, digit == i)
    mean = apply(s[-17], 2, median)
    drawn(s[1,1:16], col = "gray90")
    for (j in 2:nrow(s)) drawn(s[j,1:16], col = "gray90", add = TRUE)
    drawn(mean[-17], add = TRUE)
}
```
1. Perform a *PCA* on the data, and represent the first plan, with a specific color for each number
```{r, echo = FALSE}
res = PCA(pen, quali.sup = 17, graph = FALSE)
plot(res$ind$coord, pch = 19, cex = .5, col = rainbow(10)[pen$digit + 1],
     xlim = c(-5, 4), ylim = c(-5, 5))
legend("bottom", legend = 0:9, col = rainbow(10), pch = 19, ncol = 10,
       bty = "n", cex = .75)
res2 = data.frame(res$ind$coord, digit = factor(pen$digit))
ggplot(res2, aes(Dim.1, Dim.2, color = digit)) + geom_point()
```
1. Represent each number projections on a specific plot
```{r, echo = FALSE}
par(mfrow = c(2, 5))
for (i in 0:9) {
    s = res$ind$coord[pen$digit == i,]
    par(mar = c(2, 2, 2, 0) + .1)
    plot(s, pch = 19, cex = .5, col = rainbow(10)[i + 1],
         xlim = c(-5, 4), ylim = c(-5, 5), main = i)
}
ggplot(res2, aes(Dim.1, Dim.2, color = digit)) +
    geom_point() +
    facet_wrap(~ digit) +
    guides(color = FALSE)
```


> We conclude here that we surely need to clusters data separatly for each digit, to detect if there really are different ways to write a digit (and how).

