---
title: "`pendigits` data - interactive analysis"
subtitle: Statistical Programming using `R`
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      results = 'markdown')
library(markdown)
library(shiny)
inclure <- function(fichier) {
    name = tail(strsplit(fichier, "/")[[1]], 1)
    tags$div(
        tags$ul(tags$li(tags$a(tags$code(name), href = fichier))),
        tags$pre(tags$code(includeText(fichier)),
             class = "r")
    )
}
```

Now, we want to build an application which includes all the methods we see in previous sessions. Our goal is to create a tool that permits us to analyse the data, in order to chose an as good as possible number of clusters, for each digit.

The app includes 4 panels: the description of data and one for each method (HAC, $k$-means, mixture model). The original data are also presented. Here is the *specifications* for each panel:

- **Description**: 4 sub-panels
    - Distribution of the digits (bar chart)
    - Distribution of the coordinates, according to the digits (boxplot)
    - Average digit
    - PCA projection (global plot or separated for each digit)
- For each method:
    - Left side:
        - Choice of the digit
        - Choice of the number of clusters
    - Right side:
        - Specific graph:
            - **HAC**: dendrogram
            - **kmeans**: evolution of $r^2$
            - **mclust**: evolution of the user specified criterion ($BIC$, $ICL$, $AIC$, $AIC3$)
        - For each cluster:
            - Plot of the average digit
            - PCA projection

You can run directly this application with:

```{r, eval=FALSE, echo=TRUE}
runUrl("http://fxjollois.github.io/stat-prog-R/pendigits.zip")
```


Here are the two files for this application:

```{r}
inclure("stat-prog-R/pendigits/ui.R")
inclure("stat-prog-R/pendigits/server.R")
```


In `server.R`, we **load** a specific file, named a `RData` file. This file contains the results of the computation done in a third file (`code.R`, presented below). It allows us to pre-compute results, instead of online computing.

```{r}
inclure("stat-prog-R/pendigits/code.R")
```
